---
title: Controlling variability and maximizing effect
teaching: 20
exercises: 10
questions:
- "?"
- "?"
objectives:
- ""
- ""
keypoints:
- "Define controllable (extraneous) and uncontrollable variability in your experiment."
- "Control extraneous variance."
- "Avoid designing a complex experiment prematurely."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("07-")
```

## Controlling variability and maximizing effect
Kerlinger (1986) conceptualized experimental design as variance control [1]. The purpose of scientific research is to maximize variance explained (between-group variability, or signal) and minimize error variance (within group variability, or noise). Actual error variance should be reduced to uncontrollable variance by controlling extraneous variance.  
 
All experimental observations are a combination of signal, the true effect of a variable on an outcome, and noise, the random error inherent in your experimental technique. When designing and analyzing experiments, the goal is to maximize the signal-to-noise ratio so that you can draw accurate conclusions.
 
## Example 2 (incomplete)
from http://www.3rs-reduction.co.uk/html/7__controlling_variability.html
7. Controlling variability

```{r, example_2_table}
library(dplyr)
cgdpheno3 <- read.csv(file = "../data/CGDpheno3.csv", stringsAsFactors = FALSE)
cc_founder_phenos <- cgdpheno3 %>%
  filter(strain %in% c("A/J", "C57BL/6J", "129S1/SvImJ", "NOD/ShiLtJ",
                       "NZO/HlLtJ", "CAST/EiJ", "PWK/PhJ", "WSB/EiJ"))
cc_founder_phenos %>% group_by(strain) %>% 
  summarize(n(), mean(TG), sd(TG))
```