---
title: "Completely randomized, randomized block, and Latin square designs"
teaching: 0
exercises: 0
questions:
- "What are my experimental units?"
- "How will treatments be assigned?"
- "How will treatments be blinded?"
- "How will outcomes be measured?"
objectives:
- ""
- ""
keypoints:
- "."
- "."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
```

```{r, message=FALSE}
library(downloader)
url <- "https://raw.githubusercontent.com/smcclatchy/dals-inference/gh-pages/data/bodyWeights.csv"
filename <- "bodyWeights.csv"
if(!file.exists(filename)) download(url,destfile=filename)
```

```{r, message=FALSE}
library(dplyr)

# Read in DO850 body weight data.
dat <- read.csv("bodyWeights.csv")  
```
> ## Challenge 1
> Use this subset of data containing 20 males and 20 females and
> their baseline body weights to randomize to two different diets: high fat and regular chow.
> `subset <- dat[dat$Sample[c(51:70, 475:494)],c("Sample", "Sex", "BW.3")]`
> 1). Perform a complete randomization.  
> 2). Perform a balanced randomization. 
> 3). Check the sex ratio and difference in body weights.
> 4). Share the mean body weight for each group on the course etherpad.
>
> > ## Solution to Challenge 1
> > This requires generation of random numbers. If diets are assigned in order, sample ID will be confounded with body weight.
> > 1). 
> > 2).
> > 3).
> > 4).
> {: .solution}
{: .challenge}
