---
title: "Completely randomized, randomized block, and Latin square designs"
teaching: 0
exercises: 0
questions:
- "What are my experimental units?"
- "How will treatments be assigned?"
- "How will treatments be blinded?"
- "How will outcomes be measured?"
objectives:
- ""
- ""
keypoints:
- "."
- "."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
```

```{r, message=FALSE}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- "mice_pheno.csv"
if(!file.exists(filename)) download(url,destfile=filename)
```

```{r, message=FALSE}
library(dplyr)
dat <- read.csv("mice_pheno.csv") #Previously downloaded 

controlPopulation <- filter(dat,Sex == "F" & Diet == "chow") %>%  
  select(Bodyweight) %>% unlist

hfPopulation <- filter(dat,Sex == "F" & Diet == "hf") %>%  
  select(Bodyweight) %>% unlist

mu_hf <- mean(hfPopulation)
mu_control <- mean(controlPopulation)
print(mu_hf - mu_control)
print((mu_hf - mu_control)/mu_control * 100) #percent increase
```
> ## Challenge 1
> Use this subset of data containing 20 males and 20 females and
> their baseline body weights to randomize to two different diets: high fat and regular chow.
> 1). Perform a complete randomization.  
> 2). Perform a balanced randomization. 
> 3). Check the sex ratio and difference in body weights.
> 4). Share the mean body weight for each group on the course etherpad.
>
> > ## Solution to Challenge 1
> > This requires generation of random numbers.
> > 1). This requires generation of random numbers.
> > This requires generation of random numbers.
> > This requires generation of random numbers.
> {: .solution}
{: .challenge}
